Formalization of Typed Lua

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: rewrite of the explanation on what happens with closed-over variables, simplification of some rules

We also need to make sure to close all \emph{unique} and \emph{open}
table types before we type check a nested scope.
To do that, our type system uses some auxiliary functions to change
the type of variables before type checking a nested scope and
also to change the type of assigned and referenced variables after
type checking a nested scope.
The function \emph{crall} closes all \emph{unique} and \emph{open}
table types; it also restores filter types to their original types.
The function \emph{crset} closes a given set of free assigned variables,
which is given by the function \emph{fav}, and
it also uses this set to restore filter types to their original types.
The function \emph{openset} changes from \emph{unique} to \emph{open}
a given set of referenced variables, which is given by the function \emph{rv}.

As an example,
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Void \rightarrow \Integer \times \Void =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Void$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;$}\\
& & \multicolumn{2}{l}{$b[``x"] \; {<}\String{>} = ``foo" \;;\; f(a[``x"])$}
\end{tabular}
\end{center}
does not type check because we cannot add the field
$``x"$ to $b$, as its type is closed.
The assignment $b = a$ type checks because, at that point,
$a$ and $b$ have the same type: $\{\}_{closed}$.
Their type was closed by \emph{crall} before type checking
the function body.
Their type would be restored to $\{\}_{unique}$ after type checking
the function body, but that assignment also triggers other two type changes.
First, the function \emph{fav} includes $b$ in the set of variables
that should be closed by \emph{crset}.
Then, the function \emph{rv} includes $a$ in the set of variables
that should change from \emph{unique} to \emph{open} by \emph{openset}.
After declaring $f$, $a$ has type $\{\}_{open}$ and $b$ has type $\{\}_{closed}$,
so we can refine the type of $a$, but we cannot refine the type of $b$.

Rule \textsc{T-FUNCTION} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION}\\
\dfrac{\begin{array}{c}
       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
       \env_{3} = crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))\\
       \env_{4} = openset(\env_{3}, rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{4}}
\end{array}
\]

This rule also extends the environment $\penv$, bounding the special
variable $\ret$ to the return type $S$.
Rule \textsc{T-RETURN} uses the type that is bound to $\ret$ in
$\penv$ to type check return statements:
\[
\begin{array}{c}
\mylabel{T-RETURN}\\
\dfrac{\env_{1} \vdash el:S_{1}, \env_{2} \;\;\;
       \penv(\ret) = S_{2} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1} \vdash \mathbf{return} \; el, \env_{2}}
\end{array}
\]

@@

Notice that the local variables $q$ and $r$ are declared without 
an explicit type. The rule for typing this {\bf local} statement,
{\sc T-LOCAL-UN}, introduces a new label in the projection
environment used for typing the {\bf if} statement:
\[
\begin{array}{c}
\mylabel{T-LOCAL-UN}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\
	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
	n = |\;\overline{id}\;|  
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{4} - \{\overline{id}\}}
\end{array}
\]

This first sequent above the bar uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
This rule does the work of generating a fresh projection label,
and projection types associated with it, if the last term of
the expression list has a union of tuple types as its type,
as rule {\sc T-EXPLIST-PROJ} below shows:
\[
\begin{array}{c}
\mylabel{T-EXPLIST-PROJ}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|)-1 \;\;\; x \notin \Pi\\
	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
	\end{array}}
{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times last(S), \env_{f}, (x,S)}
\end{array}
\]

@@

At this point, using $q$ or $r$ as an r-value will project
their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. But the more interesting case is inside the
branches of the {\bf if} statement. The particular uses the rule {\sc T-IF-FNIL-PROJ} below, which uses flow typing:
\[
\begin{array}{c}
\mylabel{T-IF-FNIL-PROJ}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
       \env_{4} = closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})) \\
       \env_{5} = openset(\env_{4}, rv(s_{1}) \cup rv(s_{2}))
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{5}}
\end{array}
\]

@@

A projection type in l-value position does not trigger
the projection, effectively forbidding assignment to a
variable that has a projection type. Allowing assignment
would be unsafe, as it could break the dependency among the
components of the union, as the following example shows:
\begin{verbatim}
  local q, r = idiv(a, b)
  r = "foo"
  if q then
    -- wrongly assume q and r are integers
  end
\end{verbatim}

>>>>

Lua has first-class functions, so another way of aliasing
an {\em unique} and {\em open} references is by closing
over them in a function body. In the environment used for
typing the function body, all {\em unique} and {\em open}
tables are {\em closed} with the auxiliary function $closeall$.
This is not enough, as assignment
to a closed-over variable can be unsafe. In the following
example, assume that the $g$ receives a parameter of type
$\{ ``x": \String \cup \Nil\}$:
\begin{center}
	\begin{tabular}{llll}
		\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{open}
			= \{\} \; \mathbf{in}$}\\
		& \multicolumn{3}{l}{$\mathbf{local} \; f:\Void \rightarrow \Void =$}\\
		& & \multicolumn{2}{l}{$\mathbf{fun} \; ():\Void$}\\
		& & & \multicolumn{1}{l}{$a = \{[``x"]=5\} $}\\
		& \multicolumn{3}{l}{$\mathbf{in} \; g({<}\{``x":
			 \String \cup \Nil\}{>} \; a)$}
	\end{tabular}
\end{center}

The type of $a$ on line 3 is $\{\}_{closed}$, so the assignment
would type-check. But then the function call on the last line
would also type-check, as $\{\}_{open}$ can be coerced to
$\{ ``x": \String \cup \Nil\}$. A solution is to change the
type of $a$ from {\em unique} to {\em closed} after the function
declaration. In order to do that, we use the auxiliary function
$fav$ ({\em free assigned-to variables}) to collect this
set of identifiers, and then function $closeset$ changes
the type environment to so any of these variables that
has an {\em unique} or {\em open} table type becomes {\em closed}.

Closed-over variables that are not assigned still may
change their type if they have a {\em unique} table type
and are aliased inside the function body. We use the auxiliary
function $frv$ ({\em free referenced variables}) to collect
this set, and then function $openset$ changes any variables
with a {\em unique} table type in the environment to {\em open}.

Rule {\sc T-FUNCTION} types non-variadic function declarations,
and shows how these auxiliary functions work together:
\[
\begin{array}{c}
\mylabel{T-FUNCTION}\\
\dfrac{\begin{array}{c}
	closeall(\env_{1})[\overline{id \mapsto T}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)) \\
	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
\end{array}
%\begin{array}{c}
%\mylabel{T-FUNCTION}\\
%\dfrac{\begin{array}{c}
%	crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
%	\env_{3} = crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))\\
%	\env_{4} = openset(\env_{3}, rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
%	\end{array}}
%{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{4}}
%\end{array}
\]

Notice that the environment $\env_{2}$ resulting from
typing the function body is just ignored.

The following example typechecks, and uses all the parts of {\sc T-FUNCTION}:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; a:\{\}_{unique}, b:\{\}_{unique} = \{\}, \{\} \; \mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{local} \; f:\Integer \times \Void \rightarrow \Integer \times \Void =$}\\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer):\Integer \times \Void$}\\
& & & \multicolumn{1}{l}{$b = a \;;\; \mathbf{return} \; x + 1$}\\
& \multicolumn{3}{l}{$\mathbf{in} \; a[``x"] \; {<}\Integer{>} = 1 \;;\; f(a[``x"])$}\\
%& & \multicolumn{2}{l}{$b[``x"] \; {<}\String{>} = ``foo" \;;\;
%	 f(a[``x"])$}
\end{tabular}
\end{center}

The assignment $b=a$ typechecks because, at that point, 
both $b$ and $a$ have type $\{\}_{closed}$ due to $closeall$.
The set $fav$ is $\{b\}$, and the set $frv$ is $\{a\}$.
This causes the type of $b$ to also be $\{\}_{closed}$
in the body of the inner {\bf local} statement, due to $closeset$, and the type of $a$ to be $\{\}_{open}$ due
to $openset$. Being $\{\}_{open}$, adding a new field
is allowed, changing the type of $a$ to $\{``x": \Integer\}_{open}$ after the semicolon. After the call to
$f$, the variable $b$ is an alias to $a$ but this is not
an issue, as $\{\}_{closed}$ is a supertype of $\{``x": \Integer\}_{open}$.

@@

Notice that the local variables $q$ and $r$ are declared without 
an explicit type. The rule for typing this {\bf local} statement,
{\sc T-LOCAL-UN}, introduces a new label in the projection
environment used for typing the {\bf if} statement:
\[
\begin{array}{c}
\mylabel{T-LOCAL-UN}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:T_1 \times \ldots T_n \times P, \env_{2}, (x,S)\\
	\env_{3} = \env_{2}[id_{1} \mapsto T_1, \ldots, id_{n} \mapsto T_n]\\
	\env_{3}, \penv[x \mapsto S] \vdash s, \env_{4} \;\;\;
	n = |\;\overline{id}\;|  
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, (\env_{4}-\{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\]

The first judgement above the bar uses the auxiliary typing relation $\env_{1}, \penv \vdash el : S_{1}, \env_{2}, (x,S_{2})$.
This rule does the work of generating a fresh projection label,
and projection types associated with it, if the last term of
the expression list has a union of tuple types as its type,
as rule {\sc T-EXPLIST-PROJ} below shows:
\[
\begin{array}{c}
\mylabel{T-EXPLIST-PROJ}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
	\env_{1}, \penv \vdash me:S, \env_{n+2} \;\;\; n=|\overline{e}|\\
	S = S_1 \sqcup S_2 \;\;\;m = max(|S_1|, |S_2|)-1 \;\;\; x \notin \Pi\\
	\env_{f} = merge(\env_{1}, ..., \env_{n+2})
	\end{array}}
{\env_{1}, \penv \vdash \overline{e},me:}\\ {T_{1} \times ... \times T_{n} \times \pi_{1}^{x} \times ... \times \pi_{m}^{x} \times last(S), \env_{f}, (x,S)}
\end{array}
\]

@@

At this point, using $q$ or $r$ as an r-value will project
their types into either $\Integer \cup \Nil$ or $\Integer \cup \String$. But the more interesting case is inside the
branches of the {\bf if} statement. The particular uses the rule {\sc T-IF-FNIL-PROJ} below, which uses flow typing:
\[
%\begin{array}{c}
%\mylabel{T-IF-FNIL-PROJ}\\
%\dfrac{\begin{array}{c}
%       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
%       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
%       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
%       \env_{4} = closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})) \\
%       \env_{5} = openset(\env_{4}, rv(s_{1}) \cup rv(s_{2}))
%      \end{array}}
%      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{5}}
%\end{array}
\begin{array}{c}
\mylabel{T-IF-FNIL-PROJ}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
       \env_{1}, \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
       \env_{1}, \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
       \env_{4} = join(\env_{2}, \env_{3}) 
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\]

@@

Assignment to an l-value with a projection type is forbidden. Allowing assignment
would be unsafe, as it could break the dependency among the
components of the union, as the following example shows:
\begin{verbatim}
  local q, r = idiv(a, b)
  r = "foo"
  if q then
    -- wrongly assume q and r are integers
  end
\end{verbatim}
